---
title: 'Airline On-Time Performance'
subtitle: "Final Data Analysis"
author: "Alex Czerkawski"
output: beamer_presentation
fontsize: 10pt
---

## Review of scope

- Identifying flights at high risk of delay months in advance is a key challenge for United Airlines' network planning team
- Models will be built using historical flight data to predict the probability of departure delays at ORD (Chicago O'Hare International Airport) and PIT (Pittsburgh International Airport)
- Ideally, these models will allow planners to estimate operational impact to make better decisions
- For the scope of this project, we'll take United Airlines data between October 2024 and September 2025

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Use this block of code to import the csv
library(tidyverse)
library(caret)
library(lattice)
library(pls)
library(kableExtra)

# Import datasets
ord <- read.csv("ord.csv")

pit <- read.csv("pit.csv")

ord$Day.of.Week <- factor(ord$Day.of.Week, 
                           levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
ord$is_delayed  <- factor(ord$is_delayed, 
                           levels = c("not_delayed", "short_delay", "long_delay"))

pit$Day.of.Week <- factor(pit$Day.of.Week, 
                           levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
pit$is_delayed  <- factor(pit$is_delayed, 
                           levels = c("not_delayed", "short_delay", "long_delay"))

ordX <- ord %>% select(-is_delayed, -Dep, -Report_Date, -window_start, -window_end)
pitX <- pit %>% select(-is_delayed, -Dep, -Report_Date, -window_start, -window_end)

ord_dummies <- dummyVars(~ ., data = ordX)
ordX <- predict(ord_dummies, newdata = ordX) %>% as.data.frame()

pit_dummies <- dummyVars(~ ., data = pitX)
pitX <- predict(pit_dummies, newdata = pitX) %>% as.data.frame()
```

## Bank Background

- "Banks" are a schedule structure tool that allow airlines to provide better connectivity across a network
- Between the red lines is an example of an arrival bank (few departures)
- Between the blue lines is an example of a departure bank (many departures)
- These alternate throughout the day at hub airports

## Bank Structure at ORD

```{r,echo=FALSE, warning=FALSE}
set.seed(100)
ggplot(ord %>% slice_sample(n=800), aes(x=Scheduled.Departure.Time / 60, y=Report_Month, color = is_delayed)) + geom_point(size=1) + 
  labs(
    x = "Scheduled Departure Time (Hours)",
    y = "Month"
    ) +
  geom_vline(xintercept = c(17.5, 19), 
             color      = "blue", 
             linetype   = "dashed",
             linewidth  = 0.5) +
  geom_vline(xintercept = c(11.5, 12.5), 
             color      = "red", 
             linetype   = "dashed",
             linewidth  = 0.5) +
  scale_color_manual(values = c(
    "not_delayed" = "darkgreen",
    "short_delay" = "yellow",
    "long_delay"  = "red"
  ))
```

## Lack of Bank Structure at PIT

```{r,echo=FALSE, warning=FALSE}
ggplot(pit, aes(x=Scheduled.Departure.Time / 60, y=Report_Month, color = is_delayed)) + geom_point(size=1) + 
  labs(
    x = "Scheduled Departure Time (Hours)",
    y = "Month"
    ) +
  scale_color_manual(values = c(
    "not_delayed" = "darkgreen",
    "short_delay" = "yellow",
    "long_delay"  = "red"
  ))
```

## Delay rate by Day of Week at ORD

```{r,echo=FALSE,warning=FALSE}
ggplot(ord, aes(x = Day.of.Week, fill = factor(is_delayed, levels = c("long_delay", "short_delay", "not_delayed")))) +
  geom_bar(position = "fill") +
  labs(
    x = "Day of the Week",
    y = "Delay Rate",
    fill = "is_delayed"
  ) +
  scale_fill_manual(values = c(
    "not_delayed" = "darkgreen",
    "short_delay" = "yellow",
    "long_delay"  = "red"
  ))
```

## Delay rate by Day of Week at PIT

```{r,echo=FALSE,warning=FALSE}
ggplot(pit, aes(x = Day.of.Week, fill = factor(is_delayed, levels = c("long_delay", "short_delay", "not_delayed")))) +
  geom_bar(position = "fill") +
  labs(
    x = "Day of the Week",
    y = "Delay Rate",
    fill = "is_delayed"
  ) +
  scale_fill_manual(values = c(
    "not_delayed" = "darkgreen",
    "short_delay" = "yellow",
    "long_delay"  = "red"
  ))
```

## Model Fitting

- ORD Breakdown: On-Time = 60.8%, Short Delay = 24.9%, Long Delay = 14.3%
- PIT Breakdown: On-Time = 78.2%, Short Delay = 13.3%, Long Delay = 8.5%
- Predictors include Scheduled Departure Time, Scheduled Arrival Time, Scheduled Block Time, Day of the Week, Month, and Scheduled flights in 30 minutes before scheduled departure time (NEW)
- Three models will be fit for ORD and PIT: Logistic Regression, Random Forest, and Boosting
- Models will be evaluated on calibration and bias

```{r,echo=FALSE, warning=FALSE}
PIT_logisticReg <- readRDS("PITLogisticRegression.rds")
ORD_logisticReg <- readRDS("ORDLogisticRegression.rds")
PIT_rfTune <- readRDS("PITRandomForest.rds")
ORD_rfTune <- readRDS("ORDRandomForest.rds")
PIT_gbmFit1 <- readRDS("PITBoosting.rds")
ORD_gbmFit1 <- readRDS("ORDBoosting.rds")
```

## Model Calibration

```{r, echo=FALSE,message=FALSE,warning=FALSE}
# ORD predictions
ord_lr_prob <- predict(ORD_logisticReg, ordX, type = "prob")
ord_rf_prob <- predict(ORD_rfTune, ordX, type = "prob")
ord_boost_prob <- predict(ORD_gbmFit1, ordX, type = "prob")

ord$lr_short <- ord_lr_prob[,"short_delay"]
ord$lr_long <- ord_lr_prob[,"long_delay"]
ord$rf_short <- ord_rf_prob[,"short_delay"]
ord$rf_long <- ord_rf_prob[,"long_delay"]
ord$boost_short <- ord_boost_prob[,"short_delay"]
ord$boost_long <- ord_boost_prob[,"long_delay"]

# PIT predictions
pit_lr_prob <- predict(PIT_logisticReg, pitX, type = "prob")
pit_rf_prob <- predict(PIT_rfTune, pitX, type = "prob")
pit_boost_prob <- predict(PIT_gbmFit1, pitX, type = "prob")

pit$lr_short <- pit_lr_prob[,"short_delay"]
pit$lr_long <- pit_lr_prob[,"long_delay"]
pit$rf_short <- pit_rf_prob[,"short_delay"]
pit$rf_long <- pit_rf_prob[,"long_delay"]
pit$boost_short <- pit_boost_prob[,"short_delay"]
pit$boost_long <- pit_boost_prob[,"long_delay"]
```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
# Used claude to create this though it was based on a simpler calibration plot that I made

# Create binary factors for each class
ord$is_short_delay <- factor(ifelse(ord$is_delayed == "short_delay", "short_delay", "other"),
                              levels = c("short_delay", "other"))

ord$is_long_delay  <- factor(ifelse(ord$is_delayed == "long_delay", "long_delay", "other"),
                              levels = c("long_delay", "other"))

pit$is_short_delay <- factor(ifelse(pit$is_delayed == "short_delay", "short_delay", "other"),
                              levels = c("short_delay", "other"))

pit$is_long_delay  <- factor(ifelse(pit$is_delayed == "long_delay", "long_delay", "other"),
                              levels = c("long_delay", "other"))

# ORD - short delay
calData_ord_short <- calibration(is_short_delay ~ lr_short + rf_short + boost_short, 
                                  data = ord, cuts = 10)
cal_ord_short <- ggplot(calData_ord_short$data) +
  geom_point(aes(midpoint, Percent, color = calibModelVar)) +
  geom_line(aes(midpoint, Percent, color = calibModelVar)) +
  geom_abline() +
  labs(title = "ORD - Short Delay Calibration",
       x = "Predicted probability", y = "Observed event probability", color = "Classifier")

# ORD - long delay
calData_ord_long <- calibration(is_long_delay ~ lr_long + rf_long + boost_long, 
                                 data = ord, cuts = 10)
cal_ord_long <- ggplot(calData_ord_long$data) +
  geom_point(aes(midpoint, Percent, color = calibModelVar)) +
  geom_line(aes(midpoint, Percent, color = calibModelVar)) +
  geom_abline() +
  labs(title = "ORD - Long Delay Calibration",
       x = "Predicted probability", y = "Observed event probability", color = "Classifier")

# PIT - short delay
calData_pit_short <- calibration(is_short_delay ~ lr_short + rf_short + boost_short, 
                                  data = pit, cuts = 10)
cal_pit_short <- ggplot(calData_pit_short$data) +
  geom_point(aes(midpoint, Percent, color = calibModelVar)) +
  geom_line(aes(midpoint, Percent, color = calibModelVar)) +
  geom_abline() +
  labs(title = "PIT - Short Delay Calibration",
       x = "Predicted probability", y = "Observed event probability", color = "Classifier")

# PIT - long delay
calData_pit_long <- calibration(is_long_delay ~ lr_long + rf_long + boost_long, 
                                 data = pit, cuts = 10)
cal_pit_long <- ggplot(calData_pit_long$data) +
  geom_point(aes(midpoint, Percent, color = calibModelVar)) +
  geom_line(aes(midpoint, Percent, color = calibModelVar)) +
  geom_abline() +
  labs(title = "PIT - Long Delay Calibration",
       x = "Predicted probability", y = "Observed event probability", color = "Classifier")

# Display all 4 together
library(gridExtra)
grid.arrange(cal_ord_short, cal_ord_long, cal_pit_short, cal_pit_long, ncol = 2)
```

## Model Calibration Continued

```{r, echo=FALSE,message=FALSE,warning=FALSE}
# Used claude to create these tables
library(kableExtra)

make_cal_table <- function(calData, caption) {
  calData$data %>%
    select(calibModelVar, bin, Percent, Count) %>%
    pivot_wider(
      names_from  = calibModelVar,
      values_from = c(Percent, Count)
    ) %>%
    arrange(bin) %>%
    select(bin,
           Percent_lr_short, Count_lr_short,
           Percent_rf_short, Count_rf_short,
           Percent_boost_short, Count_boost_short) %>%
    rename(
      Bin                 = bin,
      `Pct`    = Percent_lr_short,
      `N`          = Count_lr_short,
      `Pct `    = Percent_rf_short,
      `N `          = Count_rf_short,
      `Pct  ` = Percent_boost_short,
      `N  `       = Count_boost_short
    ) %>%
    kable(caption  = caption,
          digits   = 3,
          booktabs = TRUE) %>%
    kable_styling(latex_options = c("striped", "hold_position")) %>%
    add_header_above(c(" " = 1,
                       "Logistic Regression" = 2,
                       "Random Forest"       = 2,
                       "Boosting"            = 2))
}

# For long delay tables, adjust column names in select/rename accordingly
make_cal_table_long <- function(calData, caption) {
  calData$data %>%
    select(calibModelVar, bin, Percent, Count) %>%
    pivot_wider(
      names_from  = calibModelVar,
      values_from = c(Percent, Count)
    ) %>%
    arrange(bin) %>%
    select(bin,
           Percent_lr_long, Count_lr_long,
           Percent_rf_long, Count_rf_long,
           Percent_boost_long, Count_boost_long) %>%
    rename(
      Bin                 = bin,
      `Pct`    = Percent_lr_long,
      `N`          = Count_lr_long,
      `Pct `    = Percent_rf_long,
      `N `          = Count_rf_long,
      `Pct  ` = Percent_boost_long,
      `N  `       = Count_boost_long
    ) %>%
    kable(caption  = caption,
          digits   = 3,
          booktabs = TRUE) %>%
    kable_styling(latex_options = c("striped", "hold_position")) %>%
    add_header_above(c(" " = 1,
                       "Logistic Regression" = 2,
                       "Random Forest"       = 2,
                       "Boosting"            = 2))
}

make_cal_table(cal_ord_short, "ORD - Short Delay Calibration Summary")
```

## Model Calibration Continued

```{r, echo=FALSE,message=FALSE,warning=FALSE}
make_cal_table_long(cal_ord_long,  "ORD - Long Delay Calibration Summary")
```

## Model Calibration Continued

```{r, echo=FALSE,message=FALSE,warning=FALSE}
make_cal_table(calData_pit_short, "PIT - Short Delay Calibration Summary")
```

## Model Calibration Continued

```{r, echo=FALSE,message=FALSE,warning=FALSE}
make_cal_table_long(calData_pit_long,  "PIT - Long Delay Calibration Summary")
```

## Model Application Scenario

- You work on the network planning team and a complaint has come from the operations analytics team about an under performing flight
- Your goal is to find a departure time that reduces expected delay rate
  - In a real scenario decision should balance long delay rate, short delay rate, and pricing info

## Find Underperforming Flight

```{r, echo=FALSE,message=FALSE,warning=FALSE}
df <- read.csv("final_project_dataset.csv")
df <- df %>% mutate(Report_Date = as.Date(Report_Date, origin = "1899-12-30"),
                    Report_Month = month(Report_Date),
                    Scheduled.Departure.Time = (Scheduled.Departure.Time %/% 100) * 60 + (Scheduled.Departure.Time %% 100),
                    Scheduled.Arrival.Time = (Scheduled.Arrival.Time %/% 100) * 60 + (Scheduled.Arrival.Time %% 100),
                    Actual.Departure.Time = (Actual.Departure.Time %/% 100) * 60 + (Actual.Departure.Time %% 100),
                    Actual.Arrival.Time = (Actual.Arrival.Time %/% 100) * 60 + (Actual.Arrival.Time %% 100),
                    Day.of.Week = factor(Day.of.Week, 
                                          levels = 1:7, 
                                          labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                    is_delayed = factor(ifelse(Departure.Delay > 0, "delayed", "not_delayed")),
                    cancelled = ifelse(Actual.Block.Time == 0, 1, 0)
                    )
cancelled <- df %>% filter(cancelled == 1)
not_cancelled <- df %>% filter(cancelled == 0)

not_cancelled <- not_cancelled %>% filter(Dep == "ORD")
```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
ord_grouped <- not_cancelled %>% 
  group_by(Operating.flt.num, Day.of.Week, Report_Month, is_delayed) %>% 
  summarise(n = n()) %>%
  mutate(total = sum(n),
         pct = n / total) %>% filter(is_delayed == "not_delayed")
```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
scenario_flight <- not_cancelled %>% filter(Operating.flt.num == 312, Day.of.Week == "Wed", Report_Month == 7)

scenario_flight <- scenario_flight %>% select(Operating.flt.num, Arr, Report_Date, Day.of.Week, Scheduled.Departure.Time, Departure.Delay) %>% arrange(Report_Date)

mins_to_time <- function(mins) {
  hours <- mins %/% 60
  minutes <- mins %% 60
  sprintf("%02d:%02d", hours, minutes)
}

scenario_flight$Scheduled.Departure.Time <- mins_to_time(scenario_flight$Scheduled.Departure.Time)

kable(scenario_flight, col.names = c("Flt Num", "Arr Airport", "Date", "Day", "Sch Dep Time", "Delay Min"), digits = 3, booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

- Predict probabilities for short and long delays every 5 minutes and compare expected delay rates
  - Block time = 138 min (generally doesn't change much year to year)
  - 30 minute before aggregate was taken from 07/09/2025 (schedule doesn't change much year to year)

## Apply Model

```{r, echo=FALSE,message=FALSE,warning=FALSE}
test <- ordX[1:288, ] %>% select(-n_prior_30)
test$Scheduled.Departure.Time <- seq(0, 1435, by = 5)
test$Scheduled.Block.Time <- 138
test$Scheduled.Arrival.Time <- test$Scheduled.Departure.Time + 138
test$Day.of.Week.Mon <- 0
test$Day.of.Week.Tue <- 0
test$Day.of.Week.Wed <- 1
test$Day.of.Week.Thu <- 0
test$Day.of.Week.Fri <- 0
test$Day.of.Week.Sat <- 0
test$Day.of.Week.Sun <- 0
test$Report_Month <- 7
test <- test %>%
  rowwise() %>%
  mutate(n_prior_30 = sum(
    df$Report_Date == "2025-07-09" &
    df$Scheduled.Departure.Time >= Scheduled.Departure.Time - 30 &
    df$Scheduled.Departure.Time <  Scheduled.Departure.Time
  )) %>%
  ungroup()
```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
test_prob <- predict(ORD_gbmFit1, test, type = "prob")

test$short_prob <- test_prob[,"short_delay"]
test$long_prob <- test_prob[,"long_delay"]
test$total_prob <- test$short_prob + test$long_prob

ggplot(test, aes(x = Scheduled.Departure.Time / 60)) +
  geom_line(aes(y = short_prob, color = "Short")) +
  geom_line(aes(y = long_prob,  color = "Long")) +
  geom_line(aes(y = total_prob, color = "Total")) +
  geom_hline(aes(yintercept = 0.8, color = "Current Total"),  linetype = "dashed", linewidth = 0.5) +
  geom_hline(aes(yintercept = 0.4, color = "Current Long/Short"),   linetype = "dashed", linewidth = 0.5) +
  geom_vline(xintercept = 18.25, color = "black", linetype = "solid", linewidth = 0.5) +
  scale_color_manual(values = c(
    "Short" = "#00BA38",
    "Long" = "#F8766D",
    "Total" = "#619CFF",
    "Current Total" = "#619CFF",
    "Current Long/Short" = "#F8766D"
  ),
  breaks = c("Total", "Long", "Short", "Current Total", "Current Long/Short")) +
  labs(x = "Time of Day (Hours)", y = "Expected Delay Rate", color = "Delay Type")
```