---
title: 'Homework 1: Exploratory Data Analysis'
author: "Alex Czerkawski"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Dataset choice

For this assignment, we will be using the Chemical Manufacturing Process
dataset. This dataset was taken from Applied Predictive Models (APM)
Chapter 1.4.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse)
library(caret)
library(AppliedPredictiveModeling)
data(ChemicalManufacturingProcess)
df <- ChemicalManufacturingProcess
```

## Identify an outcome variable to be predicted

-   **Yield** (continuous) = Final product yield. A summary of this
    column is:

```{r,echo=FALSE}
summary(df$Yield)
```

Below is a histogram of Yield 

```{r, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(df, aes(x=Yield)) + geom_histogram() +
  labs(
    x = "Product Yield of Chemical Manufacturing Process"
    )
```

## Preprocessing

The ChemicalManufacturingProcess data frame was pre-processed by removing all rows with missing values. Further pre-processing was applied when each model was fitted.

```{r,echo=FALSE}
df <- df %>% filter(!if_any(everything(), is.na))
dfX <- df %>% select(-Yield)
```

## Linear Regression

```{r, echo=FALSE}
ctrl <- trainControl(method = "cv", number = 10)
# Linear regression
lmTune <- train(Yield ~., df,
                  method='lm',
                  trControl=ctrl,
                  preProc = c('center', 'scale'))
lmTune
```

## Lasso

```{r, echo=FALSE}
# Lasso
set.seed(100)
lassoGrid <- expand.grid(lambda = c(0), 
                      fraction = seq(.01, 1, length = 20))
lassoTune <- train(x = (dfX), y = df$Yield,
                    method = "enet",
                    tuneGrid = lassoGrid,
                    trControl = ctrl,
                    preProc = c("center", "scale")
                  )

lassoTune
```

## Ridge Regression

```{r, echo=FALSE}
# Ridge regression
set.seed(100)
ridgeGrid <- expand.grid(lambda = seq(0, .1, length = 20))
ridgeTune <- train(x = dfX, y = df$Yield,
                   method = "ridge",
                   tuneGrid = ridgeGrid,
                   trControl = ctrl,
                   preProc = c("center", "scale")
              )
ridgeTune
```

## Support Vector Machines (SVM)

```{r, echo=FALSE}
# Support Vector Machines (SVM)
set.seed(100)
svmTune <- train(
  x=dfX, y=df$Yield, 
  method = "svmRadial",
  preProc = c("center", "scale"),
  tuneLength = 8,
  epsilon = 0.01,
  trControl = ctrl
)
svmTune
```

## Neural Nets

```{r, echo=FALSE}
# Neural Nets
nnetGrid <- expand.grid(
  decay = c(5.0, 20.0, 80.0),
  size = c(1, 2, 3)
)                        
set.seed(100)
nnetTune <- train(x=dfX, y=df$Yield,
                   method = "nnet",
                   tuneGrid = nnetGrid,
                   trControl = ctrl,
                   preProc = c("center", "scale"),
                   linout = TRUE,
                   trace = FALSE
  )
nnetTune
```

## Multivariate Adaptive Regression Splines (MARS)

```{r, echo=FALSE,warning=FALSE,message=FALSE}
# Multivariate Adaptive Regression Splines (MARS)
marsGrid <- expand.grid(.degree = 1:2, .nprune = c(20, 30, 40, 50))
set.seed(100)
marsTune <- train(x=dfX, y=df$Yield,
                   method = "earth",
                   tuneGrid = marsGrid,
                   trControl = ctrl
)
marsTune
```

## K-nearest Neighbors (KNN)

```{r, echo=FALSE}
# K-nearest Neighbors (KNN)
set.seed(100)
knnTune <- train(x = dfX,
                y = df$Yield,
                preProcess = c("center", "scale"),
                method="knn", 
                trControl=ctrl,
                tuneLength=15
            )
knnTune
```

## Regression tree

```{r, echo=FALSE,warning=FALSE,message=FALSE}
# Regression tree
set.seed(100)
cartTune <- train(x = dfX, y = df$Yield,
                  method = "rpart",
                  tuneLength = 15,
                  trControl = ctrl)
cartTune
```

A bagged regression tree model was also fitted, resulting in a reduction in cross-validated
RMSE.

```{r, echo=FALSE}
#Bagged Regression tree
set.seed(100)
treebagTune <- train(x = dfX, y = df$Yield,
                      method = "treebag",
                      nbagg = 30,
                      trControl = ctrl)
treebagTune
```

## Random forests

```{r, echo=FALSE}
# Random forests
mtryGrid <- data.frame(
  mtry = floor(seq(10, ncol(dfX), 
                      length = 10))
)
set.seed(100)
rfTune <- train(x = dfX, y = df$Yield,
                 method = "rf",
                 tuneGrid = mtryGrid,
                 ntree = 100,
                 importance = TRUE,
                 trControl = ctrl)
rfTune
```

## Model comparison



## Recommendation

