---
title: 'Homework 3: Classification'
author: "IE 2064 Data Science"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Dataset choice

For this assignment, we will be using the Hepatic Injury dataset. This dataset was taken from Applied Predictive Models (APM) Chapter 1.4.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse)
library(caret)
library(AppliedPredictiveModeling)
data(hepatic)
df <- data.frame(bio, injury)
```

## Simplify injury

Below are the levels of injury after being simplified:

```{r,echo=FALSE,warning=FALSE,message=FALSE}
df <- df %>%
  mutate(injury_binary = factor(ifelse(injury == "None", "no_injury", "injury")))
unique(df$injury_binary)
```

## Identify an outcome variable to be predicted

-   **injury_binary** (categorical) = If the liver was damaged when exposed to each compound. A summary of this column is:

```{r,echo=FALSE}
summary(df$injury_binary)
```

Below is a histogram of injury_binary 

```{r, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(df, aes(x = injury_binary)) + geom_bar()
```

## Preprocessing

Since there are no missing values in the data set, the only pre-processing that needed to be done was removing predictors with zero variance. Within each train statement, appropriate pre-processing will be applied.

```{r,echo=FALSE}
zero_var_cols <- nearZeroVar(df)
df <- df[, -zero_var_cols]

library(pROC)
build_credit_roc <- function(model) {
  probs <- predict(model, df,
                   type = "prob")[, "injury"]
  return(roc(relevel(df$injury_binary, "no_injury"), probs))
}
```

## Logistic Regression

Simple logistic regression fails because there are too many predictors compared to observations.

```{r,echo=FALSE,warning=FALSE}
ctrl <- trainControl(method = "cv", number = 10)

logisticReg <- train(injury_binary ~ .,
                     data = df,
                     method = "glm",
                     trControl = ctrl,
                     preProcess = c("center", "scale"),)
logisticReg$results
```

## Linear discriminant analysis

```{r,echo=FALSE}
lda <- train(
  injury_binary ~ .,
  data = df,
  method = "lda",
  trControl = ctrl,
  preProcess = c("center", "scale", "pca")
)
lda
```

```{r,echo=FALSE}
roc_lda <- build_credit_roc(lda)
ggroc(
  roc_lda
)

auc(roc_lda)
```

## Penalized models

```{r,echo=FALSE}
set.seed(100)
glmnGrid <- expand.grid(lambda = seq(0.0, 0.02, length = 5), 
                        alpha = seq(0.0, 1, length = 5))
glmnTuned <- train(injury_binary ~ .,
                 data = df,
                 method = "glmnet",
                 tuneGrid = glmnGrid,
                 preProc = c("center", "scale"),
                 trControl = ctrl)
glmnTuned$results
```

```{r,echo=FALSE}
roc_glmnTuned <- build_credit_roc(glmnTuned)
ggroc(
  roc_glmnTuned
)

auc(roc_glmnTuned)
```

## Classification trees

```{r,echo=FALSE}
cpGrid <- data.frame(
  cp = c(0.0, 0.005, 0.01, 0.015, 0.02, 0.03, 0.04)
)
cartTune <- train(injury_binary ~ .,
               df,
               method = "rpart",
               tuneGrid = cpGrid,
               trControl = ctrl)
cartTune
```

```{r,echo=FALSE}
roc_cartTune <- build_credit_roc(cartTune)
ggroc(
  roc_cartTune
)

auc(roc_cartTune)
```

## Random forests

```{r,echo=FALSE}
mtryGrid <- data.frame(
  mtry = c(3, 5, 7, 10, 12, 15)
)
rfTune <- train(injury_binary ~ .,
               df,
               method = "rf",
               tuneGrid = mtryGrid,
               ntree = 50,
               importance = TRUE,
               trControl = ctrl)
rfTune
```

```{r,echo=FALSE}
roc_rfTune <- build_credit_roc(rfTune)
ggroc(
  roc_rfTune
)

auc(roc_rfTune)
```

## Support Vector Machines (SVM)

```{r,echo=FALSE}
svmGrid <- expand.grid(
  C = c(0.1, 0.15, 0.2, 0.25, 0.3)
)  
set.seed(100)
svmLinearTuned <- train(
  injury_binary ~., df,
  method = "svmLinear",
  preProc = c("center", "scale"),
  tuneGrid = svmGrid,
  trControl = ctrl,
  prob.model = TRUE
)
svmLinearTuned
```

```{r,echo=FALSE}
roc_svmLinearTuned <- build_credit_roc(svmLinearTuned)
ggroc(
  roc_svmLinearTuned
)

auc(roc_svmLinearTuned)
```

## Neural Nets

```{r,echo=FALSE}
set.seed(100)
nnetGrid <- expand.grid(
  decay = c(0, 1.0, 10.0),
  size = c(1, 3, 5), # number of hidden units
  bag = TRUE
)  
nnetTune <- train(injury_binary ~., df,
                   method = "avNNet",
                   tuneGrid = nnetGrid,
                   trControl = ctrl,
                   preProc = c("center", "scale"),
                   trace = FALSE,
                   maxiter = 1000
)
nnetTune
```

```{r,echo=FALSE}
roc_nnetTune <- build_credit_roc(nnetTune)
ggroc(
  roc_nnetTune
)

auc(roc_nnetTune)
```

## Non-linear discriminant analysis

```{r,echo=FALSE}
qda <- train(
  injury_binary ~ .,
  data = df,
  method = "qda",
  trControl = ctrl,
  preProcess = c("center", "scale", "pca")
)
qda
```

```{r,echo=FALSE}
roc_qda <- build_credit_roc(qda)
ggroc(
  roc_qda
)

auc(roc_qda)
```

## K-nearest Neighbors (KNN)

```{r,echo=FALSE}
knnGrid <- expand.grid(
  k = c(2, 4, 6, 8, 10, 12)
)  
set.seed(100)
knnTune = train(injury_binary ~., df,
     method = "knn",
     trControl = ctrl,
     preProc = c("center", "scale"),
     tuneGrid = knnGrid
)
knnTune
```

```{r,echo=FALSE}
roc_knnTune <- build_credit_roc(knnTune)
ggroc(
  roc_knnTune
)

auc(roc_knnTune)
```

## Naive Bayes

```{r,echo=FALSE, warning=FALSE}
set.seed(100)
naiveBayes <- train(injury_binary ~ .,
                     data = df,
                     method = "nb",
                     trControl = ctrl,
                      tuneGrid = data.frame(
                      usekernel = TRUE,
                      fL = 1,
                      adjust = 1
  ))
naiveBayes
```

```{r,echo=FALSE,warning=FALSE}
roc_naiveBayes <- build_credit_roc(naiveBayes)
ggroc(
  roc_naiveBayes
)

auc(roc_naiveBayes)
```

## Comparison



## Recommendation

