---
title: 'Homework 3: Classification'
author: "IE 2064 Data Science"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Dataset choice

For this assignment, we will be using the Hepatic Injury dataset. This dataset was taken from Applied Predictive Models (APM) Chapter 1.4.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse)
library(caret)
library(AppliedPredictiveModeling)
data(hepatic)
df <- data.frame(bio, chem, injury)
```

## Simplify injury

Below are the levels of injury after being simplified:

```{r,echo=FALSE,warning=FALSE,message=FALSE}
df <- df %>%
  mutate(injury_binary = factor(ifelse(injury == "None", "no_injury", "injury")))
unique(df$injury_binary)
```

## Identify an outcome variable to be predicted

-   **injury_binary** (categorical) = If the liver was damaged when exposed to each compound. A summary of this column is:

```{r,echo=FALSE}
summary(df$injury_binary)
```

Below is a histogram of injury_binary 

```{r, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(df, aes(x = injury_binary)) + geom_bar()
```

## Preprocessing

Since there are no missing values in the data set, there is no pre-processed that needs to be done before fitting models. Within each train statement, appropriate pre-processing will be applied.

## Linear Classification

