---
title: 'Preliminary Data Analysis'
author: "Alex Czerkawski"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Use this block of code to import the csv
library(tidyverse)
library(caret)
library(lattice)
library(pls)
df <- read.csv("final_project_dataset.csv")
df <- df %>% mutate(Report_Date = as.Date(Report_Date, origin = "1899-12-30"),
                    Report_Month = month(Report_Date),
                    Scheduled.Departure.Time = (Scheduled.Departure.Time %/% 100) * 60 + (Scheduled.Departure.Time %% 100),
                    Scheduled.Arrival.Time = (Scheduled.Arrival.Time %/% 100) * 60 + (Scheduled.Arrival.Time %% 100),
                    Actual.Departure.Time = (Actual.Departure.Time %/% 100) * 60 + (Actual.Departure.Time %% 100),
                    Actual.Arrival.Time = (Actual.Arrival.Time %/% 100) * 60 + (Actual.Arrival.Time %% 100),
                    Day.of.Week = factor(Day.of.Week, 
                                          levels = 1:7, 
                                          labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                    is_delayed = factor(ifelse(Departure.Delay > 0, "delayed", "not_delayed")),
                    cancelled = ifelse(Actual.Block.Time == 0, 1, 0)
                    )
cancelled <- df %>% filter(cancelled == 1)
not_cancelled <- df %>% filter(cancelled == 0)
```

## CoNVO statement

United Airlines operates thousands of flights every day across a large and complex network, and identifying flights at high risk of delay months in advance is a key challenge for the network planning team. Having better visibility into which flights are likely to be delayed is an important step toward building more reliable and operationally efficient schedules.

Understanding what drives flight delays will improve foresight into where and when delays are most likely to occur. Certain routes, markets, and time slots are consistently more prone to delays than others. This will help highlight routes and markets that are consistently delayed, giving planning teams a clearer picture of where the network is underperforming. With this information, better long-term scheduling decisions can be made that improve on-time performance and reduce delays across the network.

Using historical flight data, we will build a model that predicts the probability of departure delays for flights across the network. The dataset includes flight-level information on scheduled times, routes, and day of week, providing a strong foundation for identifying where delays are most likely to occur. By identifying the key factors that contribute to delays, the model will give network planning teams the information they need to adjust routes and time slots where schedule changes could improve reliability and operational performance. Flight-level predictions will allow planners to move beyond aggregate on-time statistics and evaluate specific flights and markets that may benefit from schedule adjustments. The model will give planners a way to simulate the impact of schedule changes before they are made, helping the team make more informed decisions.

Preliminary results for departures out of ORD (Chicago O'Hare International Airport) will be presented to a VP in Network Planning. ORD is a United hub while being one of the busiest and most delay-prone airports in the country, making it a natural starting point for this analysis.  If successful, the model will be run across all flights in the network on an ongoing basis, giving planning teams regular insights into where delays are expected and where schedules may need to be adjusted. The long term goal is to make this model a regular part of the network planning process to help the team build better schedules and reduce delays across the network.

## Data Visualization

```{r,echo=FALSE,warning=FALSE}
not_cancelled <- not_cancelled %>% filter(Dep == "ORD")
```

```{r,echo=FALSE,warning=FALSE,fig.height=3.5}
set.seed(100)
ggplot(not_cancelled %>% slice_sample(n=800), aes(x=Scheduled.Departure.Time / 60, y=Report_Month, color = is_delayed)) + geom_point(size=1) + 
  labs(
    x = "Scheduled Departure Time (Hours)",
    y = "Month"
    )
```

The plot above shows how delayed flights are dispersed throughout the day in different months at ORD. There tend to be more delayed flights in the afternoon compared to the morning during all seasons, and there are certain seasons that have more delays than others.

```{r,echo=FALSE,warning=FALSE,fig.height=3.5}
set.seed(10)
ggplot(not_cancelled %>% filter(is_delayed == 'delayed') %>% slice_sample(n=800), aes(x=Scheduled.Departure.Time / 60)) + 
  geom_histogram(bins = 50, color = "white") +
  labs(
    x = "Scheduled Departure Time (Hours)",
    y = "Delay Count"
    )
```

The histogram above shows counts of delays through the day at ORD. Similar to the last plot, this graph shows more delays later in the day as well as the bank structure (peak times which increase the amount of connections United is able to offer) of the United schedule at ORD.

```{r,echo=FALSE,warning=FALSE,fig.height=3.5}
set.seed(100)
ggplot(not_cancelled, aes(x = Day.of.Week, fill = is_delayed)) +
  geom_bar(position = "fill") +
  labs(
    x = "Day of the Week",
    y = "Delay Percentage"
  )
```

The plot above shows how the percentage of flights delayed changes by the day of the week. At ORD, Tuesday has the highest percentage of on-time flights followed by Wednesday, Monday, Saturday, Thursday, Friday, and Sunday.

## Model Fitting

Below is the proportion of each level of the outcome. Models will be fit with a departure airport of ORD.

```{r,echo=FALSE, warning=FALSE}
not_cancelledX <- not_cancelled %>% select(-Marketing.Airline, -Marketing.flt.num, -Operating.Airline, -Operating.flt.num, -Report_Date, -Actual.Departure.Time, -Actual.Arrival.Time, -Actual.Block.Time, -Departure.Delay, -Arrival.Delay, -Delay.Diff, -Wheels.Up, -Wheels.Down, -cancelled, -Tail.Num, -Dep, -is_delayed, -Arr)

dummies <- dummyVars(~ ., data = not_cancelledX)
not_cancelledX <- predict(dummies, newdata = not_cancelledX) %>% as.data.frame()

prop.table(table(not_cancelled$is_delayed))
```

### Logistic Regression

```{r,echo=FALSE, warning=FALSE}
# Below is the code that was used to fit the logistic regression model
# 
# ctrl <- trainControl(method = "cv", number = 4, classProbs=TRUE, summaryFunction = twoClassSummary)
# 
# set.seed(100)
# logisticReg <- train(x = not_cancelledX, y = not_cancelled$is_delayed,
#                  method = "glm",
#                  preProc = c("center", "scale"),
#                  trControl = ctrl,
#                  metric = "ROC")
# 
# saveRDS(logisticReg, "LogisticRegression.rds")
logisticReg <- readRDS("LogisticRegression.rds")
logisticReg$results
```

### Random forest

```{r,echo=FALSE}
# Below is the code that was used to fit the random forest model
# 
# set.seed(100)
# mtryGrid <- data.frame(
#   mtry = c(1, 2, 3)
# )
# rfTune <- train(x = not_cancelledX, y = not_cancelled$is_delayed,
#                method = "rf",
#                tuneGrid = mtryGrid,
#                ntree = 200,
#                importance = TRUE,
#                trControl = ctrl,
#                metric = "ROC")
# 
# saveRDS(rfTune, "RandomForest.rds")
rfTune <- readRDS("RandomForest.rds")
best_rf <- rfTune$results[which.max(rfTune$results$ROC), ]
best_rf
```

### Boosting

```{r,echo=FALSE}
# Below is the code that was used to fit the boosting model
# 
# set.seed(100)
# gbmFit1 <- train(x = not_cancelledX, y = not_cancelled$is_delayed,
#                  method = "gbm", 
#                  trControl = ctrl,
#                  ## This last option is actually one
#                  ## for gbm() that passes through
#                  verbose = FALSE,
#                  metric = "ROC")
# 
# saveRDS(gbmFit1, "Boosting.rds")
gbmFit1 <- readRDS("Boosting.rds")
best_gbm <- gbmFit1$results[which.max(gbmFit1$results$ROC), ]
best_gbm
```

## Model comparison

```{r, echo=FALSE, message=FALSE,warning=FALSE, fig.height=4.5}
library(pROC)
build_credit_roc <- function(model) {
  probs <- predict(model, not_cancelledX,
                   type = "prob")[, "delayed"]
  return(roc(relevel(not_cancelled$is_delayed, "not_delayed"), probs))
}
ggroc(list(`Logistic regression` = build_credit_roc(logisticReg),
        `Random Forest` = build_credit_roc(rfTune),   
        `Boosting` = build_credit_roc(gbmFit1)))
```

```{r, echo=FALSE, message=FALSE,warning=FALSE}
# Used claude to create this table
library(knitr)
library(kableExtra)

# Create your data
results <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "Boosting"),
  CV_ROC = c(logisticReg$results$ROC, best_rf$ROC, best_gbm$ROC),
  AUC = c(auc(build_credit_roc(logisticReg)), auc(build_credit_roc(rfTune)), auc(build_credit_roc(gbmFit1))),
  Sensitivity = c(logisticReg$results$Sens, best_rf$Sens, best_gbm$Sens),
  Specificity = c(logisticReg$results$Spec, best_rf$Spec, best_gbm$Spec)
)

# Basic nice table
kable(results, 
      caption = "Model Comparison",
      digits = 3,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

As seen in the ROC curve above, the Random Forest model has the best AUC followed by the Boosting model and then the Logistic Regression model. When looking at the cross-validated AUC (notated as CV_ROC in the table), it is clear that the Random Forest model is over fit as it performed better on the training data than the CV test data. These models result in a similar level of AUC with all three models have an AUC that indicates improvement over selecting the most common outcome.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
not_cancelled$lr <- predict(logisticReg, not_cancelledX, type = "prob")[,1]
not_cancelled$rf <- predict(rfTune, not_cancelledX, type = "prob")[,1]
not_cancelled$boost <- predict(gbmFit1, not_cancelledX, type = "prob")[,1]
```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
calData1 <- calibration(is_delayed ~ lr + rf + boost, data = not_cancelled, 
                        cuts = 10)
calibration_curve <- ggplot(calData1$data) + 
  geom_point(aes(midpoint, Percent, color=calibModelVar)) +
  geom_line(aes(midpoint, Percent, color=calibModelVar)) +
  geom_abline() +
  labs(
  title = "Calibration plot",
  x = "Predicted probability",
  y = "Observed event probability",
  color = "classifier"
  )
calibration_curve
```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
# Used claude to create this table
cal_wide <- calData1$data %>%
  select(calibModelVar, midpoint, Percent, Count) %>%
  pivot_wider(
    names_from = calibModelVar,
    values_from = c(Percent, Count)
  ) %>%
  arrange(midpoint) %>%
  select(midpoint, 
         Percent_lr, Count_lr,
         Percent_rf, Count_rf,
         Percent_boost, Count_boost) %>%
  rename(
    Bin             = midpoint,
    `LR Avg Percent`    = Percent_lr,
    `LR Count`      = Count_lr,
    `RF Avg Percent`    = Percent_rf,
    `RF Count`      = Count_rf,
    `Boost Avg Percent` = Percent_boost,
    `Boost Count`   = Count_boost
  )

kable(cal_wide,
      caption = "Calibration Summary by Model",
      digits  = 3,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  add_header_above(c(" " = 1,
                     "Logistic Regression" = 2,
                     "Random Forest" = 2,
                     "Boosting" = 2))
```

As seen in the calibration plot and table above, the Logistic Regression and Boosting models are calibrated well. The deviation from the line in the plot by the Boosting model is caused by 0 predictions made in either of the [0:0.1] or [0.9:1] bins. A similar pattern is seen with the Logistic Regression model, though the entire distribution is biased more towards the more common outcome (no delay). When looking at the Random Forest calibration plot and table, the model over predicts low likelihood outcomes and under predicts high likelihood outcomes, indicating it is very biased towards the no delay outcome.

## Plan of work

Going forward, I want to incorporate more predictors into the model. I plan to create rolling statistics such as scheduled flights in the last 30 minutes for both United and all airlines at ORD. I will also include the arrival airport as a predictor as it was excluded in these models. Ideally, these new predictors will improve the quality of the model predictions and result in better decisions on a flight level. I also want to compare how models will perform when they are fit on non-hub departures such as flights leaving PIT.