---
title: 'Airline On-Time Performance'
subtitle: "Preliminary Data Analysis"
author: "Alex Czerkawski"
output: beamer_presentation
fontsize: 10pt
---

## Review of scope

- Identifying flights at high risk of delay months in advance is a key challenge for United Airlines' network planning team
- A model will be built using historical flight data to predict the probability of departure delays at ORD (Chicago O'Hare International Airport)
- Ideally, this model will allow planners to estimate operational impact to make better decisions
- For the scope of this project, we'll take United Airlines data between October 2024 and September 2025

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Use this block of code to import the csv
library(tidyverse)
library(caret)
library(lattice)
library(pls)
df <- read.csv("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv")
df <- df %>% mutate(Report_Date = as.Date(Report_Date, origin = "1899-12-30"),
                    Report_Month = month(Report_Date),
                    Scheduled.Departure.Time = (Scheduled.Departure.Time %/% 100) * 60 + (Scheduled.Departure.Time %% 100),
                    Scheduled.Arrival.Time = (Scheduled.Arrival.Time %/% 100) * 60 + (Scheduled.Arrival.Time %% 100),
                    Actual.Departure.Time = (Actual.Departure.Time %/% 100) * 60 + (Actual.Departure.Time %% 100),
                    Actual.Arrival.Time = (Actual.Arrival.Time %/% 100) * 60 + (Actual.Arrival.Time %% 100),
                    Day.of.Week = factor(Day.of.Week, 
                                          levels = 1:7, 
                                          labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                    is_delayed = factor(ifelse(Departure.Delay > 0, "delayed", "not_delayed")),
                    cancelled = ifelse(Actual.Block.Time == 0, 1, 0)
                    )
cancelled <- df %>% filter(cancelled == 1)
not_cancelled <- df %>% filter(cancelled == 0)

not_cancelled <- not_cancelled %>% filter(Dep == "ORD")
```

## Departures by Month and Scheduled Departure Time at ORD

- Random sample of 800 flights was taken

```{r,echo=FALSE,warning=FALSE,fig.height=5.5}
set.seed(100)
ggplot(not_cancelled %>% slice_sample(n=800), aes(x=Scheduled.Departure.Time / 60, y=Report_Month, color = is_delayed)) + geom_point(size=1) + 
  labs(
    x = "Scheduled Departure Time (Hours)",
    y = "Month"
    ) +
  geom_vline(xintercept = c(17.5, 19), 
             color      = "blue", 
             linetype   = "dashed",
             linewidth  = 0.5) +
  geom_vline(xintercept = c(11.5, 12.5), 
             color      = "red", 
             linetype   = "dashed",
             linewidth  = 0.5)
```

## Delay count by Scheduled Departure Time at ORD

```{r, echo=FALSE,warning=FALSE}
set.seed(10)
ggplot(not_cancelled %>% filter(is_delayed == 'delayed') %>% slice_sample(n=800), aes(x=Scheduled.Departure.Time / 60)) + 
  geom_histogram(bins = 50, color = "white") +
  labs(
    x = "Scheduled Departure Time (Hours)",
    y = "Delay Count"
    )
```

## Delay rate by Day of Week

```{r,echo=FALSE,warning=FALSE}
set.seed(100)
ggplot(not_cancelled, aes(x = Day.of.Week, fill = is_delayed)) +
  geom_bar(position = "fill") +
  labs(
    x = "Day of the Week",
    y = "Delay Rate"
  )
```

## Model Fitting

- % flights on-time = 60.8% (majority class)
- Predictors include Scheduled Departure Time, Scheduled Arrival Time, Scheduled Block Time, Day of the Week, and Month
- Three models will be fit, Logistic Regression, Random Forest, and Boosting
- Models will be evaluated on AUC and calibration
- Accuracy and kappa aren't as relevant since the goal is predicting an accurate delay rate, not classification

```{r,echo=FALSE, warning=FALSE}
not_cancelledX <- not_cancelled %>% select(-Marketing.Airline, -Marketing.flt.num, -Operating.Airline, -Operating.flt.num, -Report_Date, -Actual.Departure.Time, -Actual.Arrival.Time, -Actual.Block.Time, -Departure.Delay, -Arrival.Delay, -Delay.Diff, -Wheels.Up, -Wheels.Down, -cancelled, -Tail.Num, -Dep, -is_delayed, -Arr)

dummies <- dummyVars(~ ., data = not_cancelledX)
not_cancelledX <- predict(dummies, newdata = not_cancelledX) %>% as.data.frame()
```

```{r,echo=FALSE, warning=FALSE}
# Below is the code that was used to fit the logistic regression model
# 
# ctrl <- trainControl(method = "cv", number = 4, classProbs=TRUE, summaryFunction = twoClassSummary)
# 
# set.seed(100)
# logisticReg <- train(x = not_cancelledX, y = not_cancelled$is_delayed,
#                  method = "glm",
#                  preProc = c("center", "scale"),
#                  trControl = ctrl,
#                  metric = "ROC")
# 
# saveRDS(logisticReg, "LogisticRegression.rds")
logisticReg <- readRDS("LogisticRegression.rds")
```

```{r,echo=FALSE}
# Below is the code that was used to fit the random forest model
# 
# set.seed(100)
# mtryGrid <- data.frame(
#   mtry = c(1, 2, 3)
# )
# rfTune <- train(x = not_cancelledX, y = not_cancelled$is_delayed,
#                method = "rf",
#                tuneGrid = mtryGrid,
#                ntree = 200,
#                importance = TRUE,
#                trControl = ctrl,
#                metric = "ROC")
# 
# saveRDS(rfTune, "RandomForest.rds")
rfTune <- readRDS("RandomForest.rds")
best_rf <- rfTune$results[which.max(rfTune$results$ROC), ]
```

```{r,echo=FALSE}
# Below is the code that was used to fit the boosting model
# 
# set.seed(100)
# gbmFit1 <- train(x = not_cancelledX, y = not_cancelled$is_delayed,
#                  method = "gbm", 
#                  trControl = ctrl,
#                  ## This last option is actually one
#                  ## for gbm() that passes through
#                  verbose = FALSE,
#                  metric = "ROC")
# 
# saveRDS(gbmFit1, "Boosting.rds")
gbmFit1 <- readRDS("Boosting.rds")
best_gbm <- gbmFit1$results[which.max(gbmFit1$results$ROC), ]
```

## Model comparison

```{r, echo=FALSE, message=FALSE,warning=FALSE, fig.height=4.5}
# Used claude to create this table
library(knitr)
library(kableExtra)
library(pROC)

build_credit_roc <- function(model) {
  probs <- predict(model, not_cancelledX,
                   type = "prob")[, "delayed"]
  return(roc(relevel(not_cancelled$is_delayed, "not_delayed"), probs))
}

ggroc(list(`Logistic regression` = build_credit_roc(logisticReg),
        `Random Forest` = build_credit_roc(rfTune),   
        `Boosting` = build_credit_roc(gbmFit1)))
```

```{r, echo=FALSE, message=FALSE,warning=FALSE}
# Create your data
results <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "Boosting"),
  CV_ROC = c(logisticReg$results$ROC, best_rf$ROC, best_gbm$ROC),
  AUC = c(auc(build_credit_roc(logisticReg)), auc(build_credit_roc(rfTune)), auc(build_credit_roc(gbmFit1))),
  Sensitivity = c(logisticReg$results$Sens, best_rf$Sens, best_gbm$Sens),
  Specificity = c(logisticReg$results$Spec, best_rf$Spec, best_gbm$Spec)
)

# Basic nice table
kable(results, 
      digits = 3,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

## Model calibration

```{r, echo=FALSE,message=FALSE,warning=FALSE}
not_cancelled$lr <- predict(logisticReg, not_cancelledX, type = "prob")[,1]
not_cancelled$rf <- predict(rfTune, not_cancelledX, type = "prob")[,1]
not_cancelled$boost <- predict(gbmFit1, not_cancelledX, type = "prob")[,1]

calData1 <- calibration(is_delayed ~ lr + rf + boost, data = not_cancelled, 
                        cuts = 10)
```

```{r, echo=FALSE,message=FALSE,warning=FALSE}
calibration_curve <- ggplot(calData1$data) + 
  geom_point(aes(midpoint, Percent, color=calibModelVar)) +
  geom_line(aes(midpoint, Percent, color=calibModelVar)) +
  geom_abline() +
  labs(
  title = "Calibration plot",
  x = "Predicted probability",
  y = "Observed event probability",
  color = "classifier"
  )
calibration_curve
```

## Model calibration continued

```{r, echo=FALSE,message=FALSE,warning=FALSE}
# Used claude to create this table
cal_wide <- calData1$data %>%
  select(calibModelVar, midpoint, Percent, Count) %>%
  pivot_wider(
    names_from = calibModelVar,
    values_from = c(Percent, Count)
  ) %>%
  arrange(midpoint) %>%
  select(midpoint, 
         Percent_lr, Count_lr,
         Percent_rf, Count_rf,
         Percent_boost, Count_boost) %>%
  rename(
    Bin    = midpoint,
    `Pct`  = Percent_lr,
    `N`    = Count_lr,
    `Pct ` = Percent_rf,
    `N `   = Count_rf,
    `Pct  `= Percent_boost,
    `N  `  = Count_boost
  )

kable(cal_wide,
      digits  = 3,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  add_header_above(c(" " = 1,
                     "Logistic Regression" = 2,
                     "Random Forest" = 2,
                     "Boosting" = 2))
```

## Next steps

- Incorporate more predictors:
  - Scheduled flights within 30 minutes of departure to capture congestion
  - Arrival airport to capture market-specific delays
- Compare to a non-hub airport departure airport like PIT