---
title: 'Exploratory Data Analysis'
author: "Alex Czerkawski"
output: beamer_presentation
fontsize: 10pt
---

## CoNVO overview

- Identifying flights at high risk of delay months in advance is a key challenge for United Airlines' network planning team
- Understanding delay drivers will highlight underperforming routes and markets to inform better scheduling decisions
- A model will be built using historical flight data to predict the probability and intensity of departure delays
- Model outputs will give the network planning team an expectation of delays so candidate schedules can be adjusted to balance an aggressive schedule with operational efficiency

## Dataset choice

- Airline Service Quality Performance 234 dataset from Bureau of Transportation Statistics
- This dataset reports every flight in a month, when it was scheduled to leave/arrive, and when it actually left/arrived
- Data is reported by each airline
- For the scope of this project, we'll take United Airlines data between October 2024 and September 2025

<https://www.bts.gov/browse-statistical-products-and-data/bts-publications/airline-service-quality-performance-234-time>.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# This was used to save dataset to a csv
#library(readxl)
#p1 <- read_excel("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/Final Project Datasets.xlsx", sheet = "Oct-May")
#p2 <- read_excel("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/Final Project Datasets.xlsx", sheet = "Jun-Sep")
#combined <- rbind(p1, p2)
#write.csv(combined, "C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv", row.names = FALSE)
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Use this block of code to import the csv
library(tidyverse)
df <- read.csv("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv")
df <- df %>% mutate(Report_Date = as.Date(Report_Date, origin = "1899-12-30"),
                    Report_Month = month(Report_Date),
                    Scheduled.Departure.Time = (Scheduled.Departure.Time %/% 100) * 60 + (Scheduled.Departure.Time %% 100),
                    Scheduled.Arrival.Time = (Scheduled.Arrival.Time %/% 100) * 60 + (Scheduled.Arrival.Time %% 100),
                    Actual.Departure.Time = (Actual.Departure.Time %/% 100) * 60 + (Actual.Departure.Time %% 100),
                    Actual.Arrival.Time = (Actual.Arrival.Time %/% 100) * 60 + (Actual.Arrival.Time %% 100),
                    Day.of.Week = factor(Day.of.Week, 
                                          levels = 1:7, 
                                          labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                    cancelled = ifelse(Actual.Block.Time == 0, 1, 0)
                    )
cancelled <- df %>% filter(cancelled == 1)
not_cancelled <- df %>% filter(cancelled == 0)
```

## Relevant fields

- Departure Airport, Arrival Airport, 

## Pairs plot

```{r, echo=FALSE}
set.seed(2)
pairs_df <- not_cancelled %>% select(Departure.Delay, Report_Month, Day.of.Week, Scheduled.Departure.Time, Scheduled.Block.Time) %>% filter(Departure.Delay < 250)
pairs(pairs_df %>% slice_sample(n=500))
```

## Histogram of Departure Delays

```{r, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(not_cancelled, aes(x = pmin(Departure.Delay, 250))) +
  geom_histogram(bins = 50, color = "white") +
  labs(x = "Departure Delay")
```

##

```{r, echo=FALSE}
set.seed(5)
ggplot(not_cancelled %>% filter(Departure.Delay < 250) %>% slice_sample(n=1000), aes(x=Scheduled.Departure.Time, y=Departure.Delay)) + geom_point(size=1) + 
  labs(
    x = "Scheduled Departure Time",
    y = "Departure Delay"
    )
```

## Causal relationships


