---
title: 'Exploratory Data Analysis'
author: "Alex Czerkawski"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Dataset choice

For this assignment, we will be using the Chemical Manufacturing Process
dataset. This dataset was taken from Applied Predictive Models (APM)
Chapter 1.4.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse)

```

## Data set fields description

**Outcomes:**

-   **Yield** (continuous) = Final product yield. A summary of this
    column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$Yield)
```

**Predictors:**

-   **BiologicalMaterial01** (Continuous) = Measurement one of the biological 
    starting material. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$BiologicalMaterial01)
```

-   **BiologicalMaterial02** (Continuous) = Measurement two of the biological 
    starting material. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$BiologicalMaterial02)
```

-   **BiologicalMaterial03** (Continuous) = Measurement three of the biological 
    starting material. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$BiologicalMaterial03)
```

-   **BiologicalMaterial04** (Continuous) = Measurement four of the biological 
    starting material. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$BiologicalMaterial04)
```

-   **BiologicalMaterial07** (Categorical) = Measurement seven of the biological 
    starting material, taking the values:

```{r,echo=FALSE}
unique(ChemicalManufacturingProcess$BiologicalMaterial07)
```

-   **ManufacturingProcess01** (Continuous) = Measurement one of the 
    manufacturing process. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$ManufacturingProcess01)
```

-   **ManufacturingProcess02** (Categorical) = Measurement two of the 
    manufacturing process. Examples of some values include:

```{r,echo=FALSE}
head(unique(ChemicalManufacturingProcess$ManufacturingProcess02))
```
There are a total of

```{r,echo=FALSE}
length(unique(ChemicalManufacturingProcess$ManufacturingProcess02))
```

values.

-   **ManufacturingProcess03** (Continuous) = Measurement three of the 
    manufacturing process. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$ManufacturingProcess03)
```

-   **ManufacturingProcess04** (Continuous) = Measurement four of the 
    manufacturing process. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$ManufacturingProcess04)
```

-   **ManufacturingProcess12** (Categorical) = Measurement twelve of the 
    manufacturing process, taking the values:

```{r,echo=FALSE}
unique(ChemicalManufacturingProcess$ManufacturingProcess12)
```

## **CoNVO statement**

Pfizer is a pharmaceutical manufacturing company. Maximizing yield in
manufacturing vaccines is important. The industrial engineering team at Pfizer
is trying to understand what stages of their manufacturing process impact
yields the most. They will use this information to make strategic decisions
about what stages of the process need upgraded equipment to improve accuracy.

Manufacturing vaccines is a very complicated process. There are many sequential
operations, many of which impact the yield of a batch. Pfizer wants to determine
which factors make the most difference. The difficulty is that there are complex
interactions between factors, and manufacturing processes differ by the material
properties.

The project will use historical data on process yields (Chemical Manufacturing
Process dataset) to explore the effects of each factor on yield. These
relationships will be visualized through several plots to find potential
relationships. If a relationship is found, it will be reported to the IE team
who will determine how to improve that process or material property for a
larger yield.

If the head of the Pfizer IE team is happy with initial results, a deeper
analysis will be conducted to determine which interactions between material 
properties and manufacturing processes have the largest impact on yields. This
analysis will be delivered to them every quarter to determine growth
opportunities when it comes to yields.

## Pairs plot

```{r, echo=FALSE}
set.seed(1)
ChemicalManufacturingProcess <- ChemicalManufacturingProcess %>%
  mutate(across(
    c(BiologicalMaterial07, ManufacturingProcess02, ManufacturingProcess12),
    as.factor
  ))
ChemManCols <- ChemicalManufacturingProcess %>% select(Yield, BiologicalMaterial01, 
                                              BiologicalMaterial02, BiologicalMaterial03,
                                              BiologicalMaterial04, BiologicalMaterial07,
                                              ManufacturingProcess01, ManufacturingProcess02,
                                              ManufacturingProcess03, ManufacturingProcess04,
                                              ManufacturingProcess12) %>% 
                                              rename(BM1 = BiologicalMaterial01, 
                                                     BM2 = BiologicalMaterial02,
                                                     BM3 = BiologicalMaterial03,
                                                     BM4 = BiologicalMaterial04,
                                                     BM7 = BiologicalMaterial07,
                                                     MP1 = ManufacturingProcess01, 
                                                     MP2 = ManufacturingProcess02,
                                                     MP3 = ManufacturingProcess03,
                                                     MP4 = ManufacturingProcess04,
                                                     MP12 = ManufacturingProcess12
                                                     )
pairs(ChemManCols %>% slice_sample(n=40))
```

## Identify an outcome and three predictor variables

The outcome variable is yield.

We will choose BiologicalMaterial02, ManufacturingProcess02, and 
ManufacturingProcess12 as they appear uncorrelated with each other and are
highly correlated with the outcome.

## Explore the variables by applying four plot types

```{r, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(ChemicalManufacturingProcess, aes(x=Yield)) + geom_histogram() +
  labs(
    x = "Product Yield of Chemical Manufacturing Process"
    )
```

The histogram above is used to visualize the distribution of Yield values. Most
fall between 37.5 and 42.5.

```{r, echo=FALSE}
ggplot(ChemicalManufacturingProcess %>% drop_na(), 
       aes(
         x=BiologicalMaterial02, 
         y=Yield)
    ) + geom_point(size=1) + labs(
    x = "Measurement of Biological Material 2",
    y = "Product Yield of Chemical Manufacturing Process"
    )
```

Generally, there is a positive correlation between Biological Material 2 and
Yield. When the measurement for Biological Material 2 is higher, Yield tends
to be higher.

```{r, echo=FALSE}
ggplot(ChemicalManufacturingProcess %>% drop_na(), 
       aes(
         x=ManufacturingProcess02, 
         y=Yield,
         color=BiologicalMaterial02)
    ) + geom_point(size=1) + labs(
    x = "Measurement of Manufacturing Process 2",
    y = "Product Yield of Chemical Manufacturing Process"
    )
```

This graph shows the interaction between Biological Material 2 and Manufacturing
Process 2 as it related to the Yield. The majority of measurements for process 2
are between 17.5 and 22.5, with about 10 at 0. The points that are non-zero 
tend to have similar yields to the zero values.

```{r, echo=FALSE}
ggplot(ChemicalManufacturingProcess %>% 
    filter(!is.na(ManufacturingProcess12)), 
       aes(
         x=ManufacturingProcess12, 
         y=Yield)
    ) + geom_boxplot() + labs(
    x = "Measurement of Manufacturing Process 12",
    y = "Product Yield of Chemical Manufacturing Process"
    )
```

When Manufacturing Process 12 is measured at 4549, the mean and right tail are 
larger than when it is measured at 0.

## Causal relationships

We see a positive relationship between biological material 2 and yield. There is
likely some causality there, though it's unclear as there's no information about 
what biological material 2 actual measures.

There doesn't seem to be much of a relationship between manufacturing process 2
and yield. Measurements of zero and all other measurements didn't seem to make a
difference in yield. There may be a slight negative relationship in the non-zero
group of measurements in manufacturing process 2. Similar to above, there is no
information about what manufacturing process 2 is measuring, which makes it
difficult to determine causality.

There seems to be a relationship between manufacturing process 12 and yield.
When manufacturing process 12 was measured at 0, the mean yield was about 2 less
than when it's measured at 4549. Again, it's difficult to determine causality
since it's unclear what the variables are measuring.