---
title: 'Exploratory Data Analysis'
author: "Alex Czerkawski"
output: beamer_presentation
fontsize: 10pt
---

## Dataset choice

For this project, we will be using the Airline Service Quality
Performance 234 dataset from the Bureau of Transportation Statistics.
The data was reported by each carrier as mandated by the Bureauu of
Transportation Statistics. The purpose is to make performance data
available to airline consumers.

A collection of this data can be found at
<https://www.bts.gov/browse-statistical-products-and-data/bts-publications/airline-service-quality-performance-234-time>.

This project will use a year's worth of United Airlines data between October 2024 and September 2025.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# This was used to save dataset to a csv
#library(readxl)
#p1 <- read_excel("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/Final Project Datasets.xlsx", sheet = "Oct-May")
#p2 <- read_excel("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/Final Project Datasets.xlsx", sheet = "Jun-Sep")
#combined <- rbind(p1, p2)
#write.csv(combined, "C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv", row.names = FALSE)
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Use this block of code to import the csv
library(tidyverse)
df <- read.csv("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv")
df <- df %>% mutate(Report_Date = as.Date(Report_Date, origin = "1899-12-30"),
                    Report_Month = month(Report_Date),
                    Scheduled.Departure.Time = (Scheduled.Departure.Time %/% 100) * 60 + (Scheduled.Departure.Time %% 100),
                    Scheduled.Arrival.Time = (Scheduled.Arrival.Time %/% 100) * 60 + (Scheduled.Arrival.Time %% 100),
                    Actual.Departure.Time = (Actual.Departure.Time %/% 100) * 60 + (Actual.Departure.Time %% 100),
                    Actual.Arrival.Time = (Actual.Arrival.Time %/% 100) * 60 + (Actual.Arrival.Time %% 100)
                    )
```

## Data set fields description

**Outcomes:**

-   **Departure.Delay** (continuous) = Difference between scheduled departure
time and actual departure time. A summary of this column is:

```{r,echo=FALSE,warning=FALSE,message=FALSE}
summary(df$Departure.Delay)
```

-   **Arrival.Delay** (continuous) = Difference between scheduled arrival
time and actual arrival time. A summary of this column is:

```{r,echo=FALSE}
summary(df$Arrival.Delay)
```

**Predictors:**

-   **Marketing.flt.num** (Categorical) = Four digit number used by United Airlines to identify a market. Some values include: 

```{r,echo=FALSE}
head(unique(df$Marketing.flt.num))
```

There are a total of

```{r,echo=FALSE}
length(unique(df$Marketing.flt.num))
```

values.

-   **Operating.Airline** (Categorical) = Airline that operates the flight. This includes United (UA) and their regional partners. Operating.Airline takes the values:

```{r,echo=FALSE}
unique(df$Operating.Airline)
```

-   **Operating.flt.num** (Categorical) = Four digit number used by the operating airline to identify a market. Some values include: 

```{r,echo=FALSE}
head(unique(df$Operating.flt.num))
```

There are a total of

```{r,echo=FALSE}
length(unique(df$Operating.flt.num))
```

values.

-   **Dep** (Categorical) = Three letter IATA code that defines the departure airport. Some values include:

```{r,echo=FALSE}
head(unique(df$Dep))
```

There are a total of

```{r,echo=FALSE}
length(unique(df$Dep))
```

values.

-   **Arr** (Categorical) = Three letter IATA code that defines the arrival airport. Some values include:

```{r,echo=FALSE}
head(unique(df$Arr))
```

There are a total of

```{r,echo=FALSE}
length(unique(df$Arr))
```

values.

-   **Report_Month** (Categorical) = Month the flight is scheduled in, taking the values:

```{r,echo=FALSE}
sort(unique(df$Report_Month))
```

-   **Day.of.Week** (Categorical) = Day of the week the flight is scheduled on, taking the values:

```{r,echo=FALSE}
sort(unique(df$Day.of.Week))
```

Note that 1 corresponds to Monday.

-   **Scheduled.Departure.Time** (Continuous) = 

```{r,echo=FALSE}
head(unique(df$Scheduled.Departure.Time))
```

-   **scheduled.Arrival.Time** (Continuous) = 

```{r,echo=FALSE}

```

-   **Scheduled.Block.Time** (Continuous) = Amount of time the flight is schedule to take, including taxi out and taxi in time. A summary of this column is:

```{r,echo=FALSE}
summary(df$Scheduled.Block.Time)
```

-   **Tail.Num** (Categorical) = Unique identifier for the aircraft. Some values include:

```{r,echo=FALSE}
head(unique(df$Tail.Num))
```

There are a total of

```{r,echo=FALSE}
length(unique(df$Tail.Num))
```

values.

## **CoNVO statement**



## Pairs plot

```{r, echo=FALSE}
set.seed(1)

```

## Identify an outcome and three predictor variables

The outcome variable is Departure.Delay.



## Explore the variables by applying four plot types

## Single plot that shows the effects of all three predictors

## Causal relationships

## Difficulties preparing the data

## Improved analysis
