---
title: 'Exploratory Data Analysis'
subtitle: "Airline On-Time Performance"
author: "Alex Czerkawski"
output: beamer_presentation
fontsize: 10pt
---

## CoNVO overview

- Identifying flights at high risk of delay months in advance is a key challenge for United Airlines' network planning team
- Understanding delay drivers will highlight underperforming routes and markets to inform better scheduling decisions
- A model will be built using historical flight data to predict the probability and intensity of departure delays
- Model outputs will give the network planning team an expectation of delays so candidate schedules can be adjusted to balance an aggressive schedule with operational efficiency

## Dataset choice

- Airline Service Quality Performance 234 dataset from Bureau of Transportation Statistics
- This dataset reports every flight in a month, when it was scheduled to leave/arrive, and when it actually left/arrived
- Data is reported by each airline
- For the scope of this project, we'll take United Airlines data between October 2024 and September 2025

<https://www.bts.gov/browse-statistical-products-and-data/bts-publications/airline-service-quality-performance-234-time>.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# This was used to save dataset to a csv
#library(readxl)
#p1 <- read_excel("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/Final Project Datasets.xlsx", sheet = "Oct-May")
#p2 <- read_excel("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/Final Project Datasets.xlsx", sheet = "Jun-Sep")
#combined <- rbind(p1, p2)
#write.csv(combined, "C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv", row.names = FALSE)
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Use this block of code to import the csv
library(tidyverse)
df <- read.csv("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv")
df <- df %>% mutate(Report_Date = as.Date(Report_Date, origin = "1899-12-30"),
                    Report_Month = month(Report_Date),
                    Scheduled.Departure.Time = (Scheduled.Departure.Time %/% 100) * 60 + (Scheduled.Departure.Time %% 100),
                    Scheduled.Arrival.Time = (Scheduled.Arrival.Time %/% 100) * 60 + (Scheduled.Arrival.Time %% 100),
                    Actual.Departure.Time = (Actual.Departure.Time %/% 100) * 60 + (Actual.Departure.Time %% 100),
                    Actual.Arrival.Time = (Actual.Arrival.Time %/% 100) * 60 + (Actual.Arrival.Time %% 100),
                    Day.of.Week = factor(Day.of.Week, 
                                          levels = 1:7, 
                                          labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                    cancelled = ifelse(Actual.Block.Time == 0, 1, 0)
                    )
cancelled <- df %>% filter(cancelled == 1)
not_cancelled <- df %>% filter(cancelled == 0)
```

## Dataset info

- Split out cancelled flights into a separate dataset
- 1,434,047 rows, each corresponding to a departed flight
- Relevant columns: Departure airport, Arrival airport, Day of week, Scheduled departure time, Scheduled arrival time, Scheduled block time, Report month, Departure delay, and Arrival delay

## Pairs plot

- Random sample of 1000 flights

```{r, echo=FALSE}
set.seed(2)
pairs_df <- not_cancelled %>% select(Departure.Delay, Report_Month, Day.of.Week, Scheduled.Departure.Time, Scheduled.Block.Time) %>% filter(Departure.Delay < 250)
pairs(pairs_df %>% slice_sample(n=1000))
```

## Histogram of departure delays

```{r, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(not_cancelled, aes(x = pmin(Departure.Delay, 250))) +
  geom_histogram(bins = 50, color = "white") +
  labs(x = "Departure Delay")
```

## Histograms of departure delays by month

```{r, echo=FALSE}
ggplot(not_cancelled, mapping = aes(x = pmin(Departure.Delay, 250))) +
  geom_histogram(bins = 50, color = "white") +
  facet_wrap(~ Report_Month) +
  labs(x = "Departure Delay")
```

## Boxplot by day of week

- Random sample of 10,000 flights
- Filtered out flights that were delayed over 250 minutes

```{r, echo=FALSE}
set.seed(5)
ggplot(not_cancelled %>% filter(Departure.Delay < 250) %>% slice_sample(n=10000)
       , aes(x=Day.of.Week, y=Departure.Delay)) + geom_boxplot() + 
  labs(
    x = "Day of Week",
    y = "Departure Delay"
    )
not_cancelled %>% filter(Departure.Delay < 250) %>% slice_sample(n=10000) %>% 
  group_by(Day.of.Week) %>% 
  summarise(p75 = quantile(Departure.Delay, 0.75),
    p95 = quantile(Departure.Delay, 0.95))
```

## Scatterplot of departure delay vs scheduled departure time at PIT

- Filtered out flights that were delayed over 100 minutes

```{r, echo=FALSE}
ggplot(not_cancelled %>% filter(Departure.Delay < 100, Dep == "PIT"), aes(x=Scheduled.Departure.Time / 60, y=Departure.Delay)) + geom_point(size=1) + 
  labs(
    x = "Scheduled Departure Time (Hours)",
    y = "Departure Delay"
    ) +
  geom_hline(yintercept = 0, color = "red")
```

## Histogram of delays by scheduled departure time at PIT

```{r, echo=FALSE}
ggplot(not_cancelled %>% filter(Departure.Delay > 0, Dep == "PIT"), 
       aes(x = Scheduled.Departure.Time / 60)) +
  geom_histogram(binwidth = 1, color = "white") +
  labs(
    x = "Scheduled Departure Time (Hours)",
    y = "Delay Count"
  )
```

## Departure delays by month, day of week, and scheduled departure time at PIT

```{r, echo=FALSE}
set.seed(5)
ggplot(not_cancelled %>% filter(Departure.Delay < 250, Dep == 'PIT') %>% slice_sample(n=5000), 
       aes(x = Scheduled.Departure.Time, y = Departure.Delay)) + 
  geom_point(size = 1) + 
  facet_grid(Day.of.Week ~ Report_Month) +
  labs(
    x = "Scheduled Departure Time",
    y = "Departure Delay",
  ) + 
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

```

## Causal relationships

```{r, echo=FALSE,warning=FALSE,message=FALSE}
# Used claude to make this plot look nice. I came up with all the relationships myself
library(ggdag)

dag <- dagify(
  Time_Banks ~ Time_of_Day,
  Congestion_on_Tarmac ~ Time_Banks,
  Long_Taxi_In ~ Congestion_on_Tarmac,
  Late_Aircraft ~ Long_Taxi_In,
  Departure_Delay ~ Late_Aircraft,
  Morning_Delays ~ Time_of_Day,
  Aircraft_Delayed_all_day ~ Morning_Delays,
  Departure_Delay ~ Aircraft_Delayed_all_day,
  Busy_Travel_Days ~ Day_of_Week,
  Days_with_more_flights ~ Busy_Travel_Days,
  Aircrafts_breaking ~ Days_with_more_flights,
  More_aggressive_schedule ~ Days_with_more_flights,
  Busy_Travel_Months ~ Month,
  Months_with_more_flights ~ Busy_Travel_Months,
  Aircrafts_breaking ~ Months_with_more_flights,
  More_aggressive_schedule ~ Months_with_more_flights,
  Departure_Delay ~ Aircrafts_breaking,
  Less_buffer ~ More_aggressive_schedule,
  Departure_Delay ~ Less_buffer,
  coords = list(
    x = c(
      Time_of_Day = 0, Time_Banks = 1, Congestion_on_Tarmac = 2, Long_Taxi_In = 3, Late_Aircraft = 4,
      Morning_Delays = 1, Aircraft_Delayed_all_day = 2,
      Day_of_Week = 0, Busy_Travel_Days = 1, Days_with_more_flights = 2, 
      Aircrafts_breaking = 3, More_aggressive_schedule = 3, Less_buffer = 4,
      Month = 0, Busy_Travel_Months = 1, Months_with_more_flights = 2,
      Departure_Delay = 5
    ),
    y = c(
      Time_of_Day = 4, Time_Banks = 4, Congestion_on_Tarmac = 4, Long_Taxi_In = 4, Late_Aircraft = 4,
      Morning_Delays = 3, Aircraft_Delayed_all_day = 3,
      Day_of_Week = 2, Busy_Travel_Days = 2, Days_with_more_flights = 2,
      Aircrafts_breaking = 2, More_aggressive_schedule = 1, Less_buffer = 1,
      Month = 0, Busy_Travel_Months = 0, Months_with_more_flights = 0,
      Departure_Delay = 2
    )
  )
)

ggdag_classic(dag, size = 2) + 
  theme_dag()
```
