---
title: 'Exploratory Data Analysis'
author: "Alex Czerkawski"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Dataset choice

For this project, we will be using the Airline Service Quality
Performance 234 dataset from the Bureau of Transportation Statistics.
The data was reported by each carrier as mandated by the Bureauu of
Transportation Statistics. The purpose is to make performance data
available to airline consumers.

A collection of this data can be found at
<https://www.bts.gov/browse-statistical-products-and-data/bts-publications/airline-service-quality-performance-234-time>.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# This was used to save dataset to a csv
library(readxl)
p1 <- read_excel("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/Final Project Datasets.xlsx", sheet = "Oct-May")
p2 <- read_excel("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/Final Project Datasets.xlsx", sheet = "Jun-Sep")
combined <- rbind(p1, p2)
write.csv(combined, "C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv", row.names = FALSE)
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Use this block of code to import the csv
df <- read.csv("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv")
```

## Data set fields description

**Outcomes:**

-   **Yield** (continuous) = Final product yield. A summary of this
    column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$Yield)
```

**Predictors:**

-   **BiologicalMaterial01** (Continuous) = Measurement one of the
    biological starting material. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$BiologicalMaterial01)
```

-   **BiologicalMaterial02** (Continuous) = Measurement two of the
    biological starting material. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$BiologicalMaterial02)
```

-   **BiologicalMaterial03** (Continuous) = Measurement three of the
    biological starting material. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$BiologicalMaterial03)
```

-   **BiologicalMaterial04** (Continuous) = Measurement four of the
    biological starting material. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$BiologicalMaterial04)
```

-   **BiologicalMaterial07** (Categorical) = Measurement seven of the
    biological starting material, taking the values:

```{r,echo=FALSE}
unique(ChemicalManufacturingProcess$BiologicalMaterial07)
```

-   **ManufacturingProcess01** (Continuous) = Measurement one of the
    manufacturing process. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$ManufacturingProcess01)
```

-   **ManufacturingProcess02** (Categorical) = Measurement two of the
    manufacturing process. Examples of some values include:

```{r,echo=FALSE}
head(unique(ChemicalManufacturingProcess$ManufacturingProcess02))
```

There are a total of

```{r,echo=FALSE}
length(unique(ChemicalManufacturingProcess$ManufacturingProcess02))
```

values.

-   **ManufacturingProcess03** (Continuous) = Measurement three of the
    manufacturing process. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$ManufacturingProcess03)
```

-   **ManufacturingProcess04** (Continuous) = Measurement four of the
    manufacturing process. A summary of this column is:

```{r,echo=FALSE}
summary(ChemicalManufacturingProcess$ManufacturingProcess04)
```

-   **ManufacturingProcess12** (Categorical) = Measurement twelve of the
    manufacturing process, taking the values:

```{r,echo=FALSE}
unique(ChemicalManufacturingProcess$ManufacturingProcess12)
```

## **CoNVO statement**

## Pairs plot

```{r, echo=FALSE}
library(tidyverse)
set.seed(1)
ChemicalManufacturingProcess <- ChemicalManufacturingProcess %>%
  mutate(across(
    c(BiologicalMaterial07, ManufacturingProcess02, ManufacturingProcess12),
    as.factor
  ))
ChemManCols <- ChemicalManufacturingProcess %>% select(Yield, BiologicalMaterial01, 
                                              BiologicalMaterial02, BiologicalMaterial03,
                                              BiologicalMaterial04, BiologicalMaterial07,
                                              ManufacturingProcess01, ManufacturingProcess02,
                                              ManufacturingProcess03, ManufacturingProcess04,
                                              ManufacturingProcess12) %>% 
                                              rename(BM1 = BiologicalMaterial01, 
                                                     BM2 = BiologicalMaterial02,
                                                     BM3 = BiologicalMaterial03,
                                                     BM4 = BiologicalMaterial04,
                                                     BM7 = BiologicalMaterial07,
                                                     MP1 = ManufacturingProcess01, 
                                                     MP2 = ManufacturingProcess02,
                                                     MP3 = ManufacturingProcess03,
                                                     MP4 = ManufacturingProcess04,
                                                     MP12 = ManufacturingProcess12
                                                     )
pairs(ChemManCols %>% slice_sample(n=40))
```

## Identify an outcome and three predictor variables

The outcome variable is yield.

We will choose BiologicalMaterial02, ManufacturingProcess02, and
ManufacturingProcess12 as they appear uncorrelated with each other and
are highly correlated with the outcome.

## Explore the variables by applying four plot types

```{r, echo=FALSE,warning=FALSE,message=FALSE}
ggplot(ChemicalManufacturingProcess, aes(x=Yield)) + geom_histogram() +
  labs(
    x = "Product Yield of Chemical Manufacturing Process"
    )
```

The histogram above is used to visualize the distribution of Yield
values. Most fall between 37.5 and 42.5.

```{r, echo=FALSE}
ggplot(ChemicalManufacturingProcess %>% drop_na(), 
       aes(
         x=BiologicalMaterial02, 
         y=Yield)
    ) + geom_point(size=1) + labs(
    x = "Measurement of Biological Material 2",
    y = "Product Yield of Chemical Manufacturing Process"
    )
```

Generally, there is a positive correlation between Biological Material 2
and Yield. When the measurement for Biological Material 2 is higher,
Yield tends to be higher.

```{r, echo=FALSE}
ggplot(ChemicalManufacturingProcess %>% drop_na(), 
       aes(
         x=ManufacturingProcess02, 
         y=Yield,
         color=BiologicalMaterial02)
    ) + geom_point(size=1) + labs(
    x = "Measurement of Manufacturing Process 2",
    y = "Product Yield of Chemical Manufacturing Process"
    )
```

This graph shows the interaction between Biological Material 2 and
Manufacturing Process 2 as it related to the Yield. The majority of
measurements for process 2 are between 17.5 and 22.5, with about 10 at
0. The points that are non-zero tend to have similar yields to the zero
values.

```{r, echo=FALSE}
ggplot(ChemicalManufacturingProcess %>% 
    filter(!is.na(ManufacturingProcess12)), 
       aes(
         x=ManufacturingProcess12, 
         y=Yield)
    ) + geom_boxplot() + labs(
    x = "Measurement of Manufacturing Process 12",
    y = "Product Yield of Chemical Manufacturing Process"
    )
```

When Manufacturing Process 12 is measured at 4549, the mean and right
tail are larger than when it is measured at 0.

## Causal relationships

## Difficulties preparing the data

## Improved analysis
