---
title: 'Preliminary Data Analysis'
author: "Alex Czerkawski"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Use this block of code to import the csv
library(tidyverse)
library(caret)
library(lattice)
library(pls)
df <- read.csv("C:/Users/aczer/OneDrive/Documents/Pitt Classwork/8 - Senior Spring/Applied Data Analytics/final_project_dataset.csv")
df <- df %>% mutate(Report_Date = as.Date(Report_Date, origin = "1899-12-30"),
                    Report_Month = month(Report_Date),
                    Scheduled.Departure.Time = (Scheduled.Departure.Time %/% 100) * 60 + (Scheduled.Departure.Time %% 100),
                    Scheduled.Arrival.Time = (Scheduled.Arrival.Time %/% 100) * 60 + (Scheduled.Arrival.Time %% 100),
                    Actual.Departure.Time = (Actual.Departure.Time %/% 100) * 60 + (Actual.Departure.Time %% 100),
                    Actual.Arrival.Time = (Actual.Arrival.Time %/% 100) * 60 + (Actual.Arrival.Time %% 100),
                    Day.of.Week = factor(Day.of.Week, 
                                          levels = 1:7, 
                                          labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                    is_delayed = factor(ifelse(Departure.Delay > 0, "delayed", "not_delayed")),
                    cancelled = ifelse(Actual.Block.Time == 0, 1, 0)
                    )
cancelled <- df %>% filter(cancelled == 1)
not_cancelled <- df %>% filter(cancelled == 0)
```

## CoNVO statement

United Airlines operates thousands of flights every day across a large and complex network, and identifying flights at high risk of delay months in advance is a key challenge for the network planning team. Having better visibility into which flights are likely to be delayed is an important step toward building more reliable and operationally efficient schedules.

Understanding what drives flight delays will improve foresight into where and when delays are most likely to occur. This will help highlight routes and markets that are consistently delayed, giving planning teams a clearer picture of where the network is underperforming. With this information, better long-term scheduling decisions can be made that improve on-time performance and reduce delays have across the network.

Using historical flight data, we will build a model that predicts both the probability and intensity of departure delays for flights across the network. By identifying the key factors that contribute to delays, the model will give network planning teams the information they need to adjust routes and time slots where schedule changes could improve reliability and operational performance.

Preliminary results will be presented to a VP in Network Planning. If successful, the model will be run across all flights in the network on an ongoing basis, giving planning teams regular insights into where delays are expected and where schedules may need to be adjusted. The long term goal is to make this model a regular part of the network planning process to help the team build better schedules and reduce delays across the network.

## Data Visualization

```{r,echo=FALSE,warning=FALSE}

```

## Model Fitting

```{r,echo=FALSE, warning=FALSE}
not_cancelledX <- not_cancelled %>% filter(Dep == "ORD") %>% select(-Marketing.Airline, -Marketing.flt.num, -Operating.Airline, -Operating.flt.num, -Report_Date, -Actual.Departure.Time, -Actual.Arrival.Time, -Actual.Block.Time, -Departure.Delay, -Arrival.Delay, -Delay.Diff, -Wheels.Up, -Wheels.Down, -cancelled, -Tail.Num, -Dep, -is_delayed, -Arr) #%>% mutate(Arr = as.factor(Arr))

dummies <- dummyVars(~ ., data = not_cancelledX)
not_cancelledX <- predict(dummies, newdata = not_cancelledX) %>% as.data.frame()

not_cancelledY <- not_cancelled %>% filter(Dep == "ORD")

prop.table(table(not_cancelledY$is_delayed))
```

### Logistic Regression

```{r,echo=FALSE, warning=FALSE}
# Below is the code that was used to fit the logistic regression model

ctrl <- trainControl(method = "cv", number = 4, classProbs=TRUE, summaryFunction = twoClassSummary)

set.seed(100)
logisticReg <- train(x = not_cancelledX, y = not_cancelledY$is_delayed,
                 method = "glm",
                 preProc = c("center", "scale"),
                 trControl = ctrl,
                 metric = "ROC")

saveRDS(logisticReg, "LogisticRegression.rds")
logisticReg <- readRDS("LogisticRegression.rds")
logisticReg
```

### Random forests

```{r,echo=FALSE}
set.seed(100)
mtryGrid <- data.frame(
  mtry = c(1, 2, 3, 4)
)
rfTune <- train(x = not_cancelledX, y = not_cancelledY$is_delayed,
               method = "rf",
               tuneGrid = mtryGrid,
               ntree = 10,
               importance = TRUE,
               trControl = ctrl,
               metric = "ROC")
rfTune
```

```{r,echo=FALSE}

```

### Boosting

```{r,echo=FALSE}
set.seed(100)
gbmFit1 <- train(x = not_cancelledX, y = not_cancelledY$is_delayed,
                 method = "gbm", 
                 trControl = ctrl,
                 ## This last option is actually one
                 ## for gbm() that passes through
                 verbose = FALSE,
                 metric = "ROC")
gbmFit1
```

## Model comparison

```{r, message=FALSE,warning=FALSE, fig.height=4.5}
library(pROC)
build_credit_roc <- function(model) {
  probs <- predict(model, not_cancelledX,
                   type = "prob")[, "delayed"]
  return(roc(relevel(not_cancelledY$is_delayed, "not_delayed"), probs))
}
ggroc(list(`Logistic regression` = build_credit_roc(logisticReg),
        `Random Forest` = build_credit_roc(rfTune),   
        `Boosting` = build_credit_roc(gbmFit1)))
```

## Model comparison



## Plan of work

